Forest canopies are better generators of biodiversity in the tropics

### 2025-08-26 manuscript.Rmd
Figure 1 caption update

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humboldt began an expedition that would ultimately canonize the role of mountains as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019]. This pattern—'Humboldt's enigma'—can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019]. The vertical nature of mountains provides more land surface per unit area than flat regions and brings distinct climates within close proximity of each, which produces rapid rates of species turnover. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. This vertical stacking of species with increasing arboreal niche opportunities has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024]. Yet taller canopies are not consistently associated with higher biodiversity across the globe [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents 'MacArthur's enigma'.

@@

```{r fig1, fig.id="fig1", fig.cap="Spatial distribution of a) canopy height, b) vertebrate richness (for all vertebrates), and c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height. d) Relationships between species richness of all vertebrates and canopy height across biomes. Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=140}

@@

*Micropoint* requires a number of additional parameters describing vegetation structure that are not available from remote sensing products, such leaf angle distribution and stomatal conductance, among others. We use *microclimf's* 'vegpfromhab()' function to estimate these parameters based on land cover type. Land cover was determined based on MODIS land cover type (MCD12Q1) version 6.1 for the year 2022.

@@

[WE SHOULD PRODUCE MODELS AT DRIEST MONTH OF THE YEAR]

>>>>

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humboldt began an expedition that would ultimately canonize the role of mountains as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019]. This pattern—'Humboldt's enigma'—can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019]. The vertical nature of mountains provides more land surface per unit area than flat regions and brings distinct climates within close proximity of each, which produces rapid rates of species turnover. Yet mountains are better generators of biodiversity in the tropics because they have higher elevation snow lines than high-latitude mountains, which increase the area of land suitable for diverse communities, longer climate gradients due to warmer lowlands, and more distinct elevational bands due to less seasonal temperature variation [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. This vertical stacking of species with increasing arboreal niche opportunities has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024]. Yet MacArthur & MacArthur [ -@macarthurBirdSpeciesDiversity1961] noted over 50 years ago that tropical forests support greater bird diversity than temperate forests with comparable profiles, and more recent studies have revealed inconsistent effects of canopy height on biodiversity across the globe [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical component to Humboldt's enigma.

@@

```{r fig1, fig.id="fig1", fig.cap="a) Relationships between species richness and canopy height across all vertebrate taxa globally and in different biomes. Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of canopy height (black) or richness as a function of the interaction between canopy height and biome. b-d) Spatial distribution of canopy height, vertebrate richness (for all vertebrates), and deviance residuals for the global model of vertebrate richness.", fig.height=100, fig.width=140}

@@

*Micropoint* requires a number of additional parameters describing vegetation structure that are not available from remote sensing products, such leaf angle distribution and stomatal conductance, among others. We use *microclimf's* 'vegpfromhab()' function to estimate these parameters based on land cover type. Land cover was determined based on Copernicus Global Dynamic Land Cover version 3 for the year 2019 [@buchhornCopernicusGlobalLand2020].

@@
